FraCaS: Temporal Analysis

### 2020-04-02 paper2020/paper.tex
revision

\usepackage[hyperref]{naaclhlt2019}
\usepackage{natbib}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{todonotes}
\usepackage{enumitem}
\newlist{lingex}{enumerate}{3}
\setlist[lingex,1]{parsep=0pt,itemsep=1pt,label=(\arabic*),resume=lingexcount}
\newcommand\onelingex[1]{\begin{lingex}\item #1 \end{lingex}}

@@

We are relying on a largely standard montegovian semantics. In brief,
sentences are interpreted as propositions, verbs and noun-phrases as
predicates.  We use type-raising for noun-phrases to support
quantifiers.

We support additionally various constructinos and phenomena present in
the testsuite, including adjectives, adverbs, nouns, verbs, anaphora,
etc. We direct the reader to our previous work for details \ifanon
(citations anonymized) \else \citet{} \fi, but the particular
treatment of such phenomena is largely independent from our treatment
of time.
%
Here we simply ignore these aspects beyond the fact that they
are handled correctly in most of the case of the FraCaS testsuite.  In
particular, non-temporal anaphoroi are resolved as they should be.

@@

In montegovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise verb to take two
additional time parameters, one corresponding with the starting time
of the action and one corresponding to its stopping time. For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). We are thus
using a simple newtonian model of time, corresponding to layman
intuition.

@@

We adjust the montegovian semantics so that the interpretation of
every category (propositions, verb phrases, etc.) takes an
\emph{optional} timespan as an additional parameter. This timespan
serves as a time reference for semantics, and we call it a
\emph{temporal context}. This context propagates through the
compositional interpretation down to lexical items with atomic
representation (verbs). By default, every interpretation passes the
temporal context down to its components without changing it. However
some key elements will act on it on nontrivial ways, which we proceed
to detail below.

\newcommand\nospan{\ensuremath{─}}
\newcommand\anyspan{\ensuremath{\_}}
The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on a timespan, we will write \anyspan{} instead

@@

The principal non-trivial manipulation of timespans are tense
markers. In our syntax, inherited from GF, tenses are represented
syntactically as an attribute of the clause.

Example:

A scandinavian won the nobel prize.

useCl past pPos (predVP (detCN (detQuant indefArt numSg) scandinavian\_CN) (complSlash (slashV2a win\_V2) (detCN (detQuant indefArt numSg) nobel\_prize\_CN)))

In our semantics we deal only with present and simple past
tenses. \jp{Is this enough for theory?} Indeed we find that FraCas does not exercise additional
specific tenses (if a more complicated tense is used, the additional
information is also carried by adverbs or adverbial phrases, in a more
specific way.)

The procedure to handle tense annotations is as follows:
\begin{itemize}
\item If the temporal context was modified by an adverb (or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$, which is a logical constant representing the current
  time.
\item Otherwise, if the tense is the present, then the temporal
  context becomes the simple $(now,now)$ interval.
\item
  If the tense is the past, then we locally quantify over a time
  interval $[t_0,t_1]$, such that $t_1 < now$.
\end{itemize}

@@

\begin{enumerate}[align=left,font=\itshape]
\item[exact adverbs] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in someway. Sometimes the restriction is an equality (``for exactly
  two hours'').

@@

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \begin{multline*}
  ⟦\text{always \(s\)}⟧(t_0,t_1) = \\ ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')
\end{multline*}
Note that, here, we do use the input interval, allowing to interpret
phrases such as ``In 1994, Itel was always on time.''.

@@

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid one \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:

\begin{enumerate}[parsep=0pt,itemsep=1pt,label=(\arabic*)]
\item Smith left at 1pm.
\item Smith went to its appointment with the lawyer.
\item Smith came back at 2pm
\item Smith left at 4pm.
\end{enumerate}
One would normally not say that there is contradiction. However if (1)
and (4) are presented in isolation, a contradiction can be flagged.

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because being in Paris is stative, we also get that
Smith was in Paris in any subinterval.

\paragraph{(un)repeatable achievements}
The principle of using unicity of action interacts well with the usual interpretation of quantifiers (and anaphora).
Indeed, using it, we can refute problem (279), as expected in the testsuite:

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, and thus we cannot deduce equality of the
intervals $t_1,t_2$ and $t_3,t_4$. In turn, the hypothesis cannot be refuted.

@@

The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank anonymous reviewers for their useful comments on an
earlier draft of the paper.

\bibliographystyle{acl_natbib}

>>>>

\usepackage[hyperref]{naaclhlt2019}
\usepackage{natbib}
\usepackage{tabularx}
\usepackage{times}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{todonotes}
\usepackage{enumitem}
\newlist{lingex}{enumerate}{3}
\setlist[lingex,1]{parsep=0pt,itemsep=1pt,label=(\arabic*),resume=lingexcount}
\newcommand\onelingex[1]{\begin{lingex}\item #1 \end{lingex}}

@@

Our interpretation is cast withing the framework of a largely standard
montegovian semantics \cite{todo}. In brief, sentences are interpreted as
propositions, verbs and noun-phrases as predicates.  We use
type-raising of noun-phrases, to support quantifiers \cite{PTQ}.

We support additionally various constructinos and phenomena present in
the testsuite, including adjectives, adverbs, nouns, verbs, anaphora,
etc. We direct the reader to our previous work for details \ifanon
(citations anonymized) \else \citet{bernardy_type_2017,bernardy_wide_2019} \fi, but the particular
treatment of such phenomena is largely independent from our treatment
of time.
%
Here we simply ignore these aspects beyond the fact that they
are handled correctly in most of the case of the FraCaS testsuite.  In
particular, non-temporal anaphoroi are resolved as they should be.

@@

In montegovian semantics, (intransitive) verbs are one-place
predicates; in types, they are functions from entities to propositions
($e \to t$).  Our basic approach is to generalise the interpretation of 
verbs, so that it takes two
additional time parameters, one corresponding with the starting time
of the action and one corresponding to its stopping time ($(time × time × e) \to t$). For example,
if John walked between $t_0$ and $t_1$, we would have:
$walk(john,t_0,t_1)$. From now on we will call an interval of time
points $[t_0,t_1]$ a timespan. Every timespan $[t_0,t_1]$ has the
property $t_0 ≤ t_1$ (it starts no later than it stops). (We are thus
using a simple newtonian model of time, corresponding to a layman
intuition.)

@@

We adjust the montegovian semantics so that the interpretation of
every category (propositions, verb phrases, etc.) takes a
\emph{temporal context} as an additional parameter, serves as a time
reference for the interpretation of all time-dependent semantics
within the phrase.  This context propagates through the compositional
interpretation down to lexical items with atomic representation
(verbs). By default, every interpretation passes the temporal context
down to its components without changing it. However some key elements
will act on it on nontrivial ways, which we proceed to detail below.

This temporal context is an \emph{optional} timespan. That is, it can
be an timespan or an explicitly unspecified context.
\newcommand\nospan{\ensuremath{─}}
\newcommand\anyspan{\ensuremath{\_}} The timespan in the context is
optional because, in certain situations, the semantics is different
depending on whether a timespan has been specified externally or not,
as we explain below. A non-present timespan will be represented as
\nospan{}. If a semantic function does not depend on the temporal
context at all, we will write \anyspan{} instead.

@@

\newcommand\varid[1]{\mathnormal{#1}}
%
The principal non-trivial manipulator of timespans are tense
markers. In our syntax, inherited from GF, tenses are represented
syntactically as an attribute of the clause. An illustration of a
past-tense clause and its interpretation follows. Notice in particular
the $past$ argument to $useCl$.

\onelingex{A scandinavian won the nobel prize.\label{ex:onelingex}
\\
{\small\begin{multline*}
\varid{useCl}\, \varid{past}\, \varid{pPos}\\ (\varid{predVP}\, (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{scandinavian\_CN})\\ (\varid{complSlash}\, (\varid{slashV2a}\, \varid{win\_V2})\\ (\varid{detCN}\, (\varid{detQuant}\, \varid{indefArt}\, \varid{numSg})\\ \varid{nobel\_prize\_CN})))
\end{multline*}
}}
In our semantics we deal only with present and simple past
tenses. \jp{Is this enough for theory?} Indeed we find that FraCas
does not exercise additional specific tenses (if a more complicated
tense is used, the additional information is also carried by adverbs
or adverbial phrases, in a more specific way.)

The procedure to handle tense annotations is as follows:
\begin{itemize}
\item If the tense is the past, and the temporal context is
  unspecified, then we locally quantify over a time interval
  $[t_0,t_1]$, such that $t_1 < now$, where $now$ is a logical
  constant representing the current timepoint. The temporal context
  then becomes this interval.
\item If the tense is the present and the temporal context is
  unspecified, then the temporal context becomes the simple
  $(now,now)$ interval.
\item If the temporal context is specified (for example due to the presence of an adverb or an
  averbial clause, such as ``before James swam''), then the tense does
  not create a new interval, but it may constrain it. Typically, a
  past tense adds the constraint that the temporal context ends before
  the timepoint $now$.
\end{itemize}

@@

\begin{enumerate}[align=left,font=\itshape]
\item[exact] For such adverbs, an exact interval is
  provided. Typical such adverbs specify in fact a single point in
  time (so the start and the end of the interval coincide).

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in some way. Sometimes the restriction is an equality, as in ``for
  exactly two hours''.

@@

  If there is no explicit time context, then ``always'' has no
  constraint on the interval, otherwise the quantified interval must
  be included in it:
  \begin{multline*}
  ⟦\text{always \(s\)}⟧(t_0,t_1) = \\ ∀t_0',t_1'. [t_0',t_1'] ⊆ [t_0,t_1], ⟦s⟧(t_0',t_1')
\end{multline*}
Note that here we \emph{do} use the input interval, resulting in a correct interpretation for
phrases such as ``In 1994, Itel was always on time.'' .

@@

Unicity of action appears to be a non-logical principle. Indeed, it is
quite possible that ``Jones left'' several times. However, it seems
that this principle is never contradicted by the testsuite. As such,
even though unicity of action is only a pragmatic rule, it can be
taken as a valid one \emph{by default}: it is only when we have a
sufficiently constrained situation that one should reject it. Consider
the following discourse:
\begin{enumerate}[parsep=0pt,itemsep=1pt,label=(\arabic*)]
\item Smith left at 1pm.
\item Smith went to its appointment with the lawyer.
\item Smith left at 4pm.
\end{enumerate}
One would normally not say that there is contradiction. However if the
middle sentence were not present, a contradiction should be
flagged. We leave such discourse analysis as future work, and simply
apply unicity of action everywhere.

@@

Indeed, from \textbf{P3} we get that Smith was in paris between May
5th and May 15th. Because being in Paris is stative, we also get that
Smith was in Paris in any subinterval. Contrary to unicity of action,
subsumption is always valid.

\paragraph{(un)repeatable achievements}
The principle of using unicity of action interacts well with the usual
interpretation of quantifiers (and anaphora).  Indeed, using it, we
can refute problem (279), as expected by the testsuite:

@@

\[\begin{array}{l}
∃x. novel(x) ∧ \\
∃t_1,t_2. [t_1,t_2] ⊆ 1991 ∧ write(smith,x,t_1,t_2) ∧ \\
∃y. novel(y) ∧ \\
∃t_3,t_4. [t_3,t_4] ⊆ 1992 ∧ write(smith,y,t_3,t_4) \\
⟶ ⊥ \\
\end{array}
\]
Our analysis does not need to treat this last case specially. Indeed,
even if $write(smith,x,.,.)$ is an activity and thus subject to
unicity of action, in (280), $x$ is quantified existentially; we have
two \emph{different} actions: $write(smith,x,t_1,t_2)$ and
$write(smith,y,t_3,t_4)$, because $x \neq y$, and thus we cannot deduce equality of the
intervals $t_1,t_2$ and $t_3,t_4$. In turn, the hypothesis cannot be refuted.

@@

\ifanon
Anonymized
\else
The research reported in this paper was supported by grant 2014-39 from the
Swedish Research Council, which funds the Centre for Linguistic Theory and
Studies in Probability (CLASP) in the Department of Philosophy, Linguistics,
and Theory of Science at the University of Gothenburg. We are grateful to
our colleagues in CLASP for helpful discussion of some of the ideas presented
here. We also thank anonymous reviewers for their useful comments on an
earlier draft of the paper.
\fi
\bibliographystyle{acl_natbib}
